Model Checking

### 2023-05-25 owen/Write Ups/03_existing_models.qmd
update the data to use normalised running average

library(tidyverse)
library(RcppRoll)
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)

@@

get_hist <- function(user, games, prev_n) {
  hist_games <- games %>% 
    filter(White == user | Black == user) %>% 
    arrange(UTCDate, UTCTime) %>% 
    mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
    select(White:BlackElo, focal_white, game_id) %>% 
    mutate(focal_result = case_when(
      (focal_white == 1 & Result == "1-0") ~ 1,
      (focal_white == 0 & Result == "0-1") ~ 1,
      (Result == "1/2-1/2") ~ 0.5,
      .default = 0
      )) %>% 
    mutate(focal_win_prop = c(cumsum(focal_result[1:(prev_n - 1)])/(1:(prev_n -1)), 
                              roll_mean(focal_result, n = prev_n)))s
  hist_games
}

@@

## then just get all the history for this
all_hist <- map_dfr(players, get_hist_opp, bullet_60, prev_n = 5) %>% 
  as_tibble() %>% 
  select(game_id:player)

@@

tidy_stan <- tidy_bullet %>% 
  left_join(all_hist, by = c("game_id", "Username" = "player")) %>% 
  arrange(UTCDate, UTCTime) %>% 
  rename(focal_result = opp_result, focal_win_prop = opp_win_prop) %>% 
  left_join(all_hist, by = c("game_id", "opponent" = "player")) %>% 
  as_tibble()

@@

## getting the overall win probabilities for each player also

player_avg <- all_hist %>% 
  group_by(player) %>% 
  summarise(overall_avg = mean(opp_result))

@@

final_data <- tidy_stan %>% 
  left_join(player_avg, by = c("Username" = "player")) %>% 
  rename(focal_avg = overall_avg) %>% 
  left_join(player_avg, by = c("opponent" = "player")) %>% 
  rename(opp_avg = overall_avg) %>% 
  group_by(Username) %>% 
  arrange(UTCDate, UTCTime) %>% 
  mutate(focal_prev_result = lag(focal_result, default = 0)) %>% 
  ungroup() %>% 
  group_by(opponent) %>% 
  arrange(UTCDate, UTCTime) %>% 
  mutate(opp_prev_result = lag(opp_result, default = 0)) %>% 
  ungroup() %>% 
  mutate(focal_id = match(Username, players),
         opp_id = match(opponent, players)) %>% 
  filter(focal_result != 0.5)

@@

@@

stan_data_all <- list(N = nrow(final_data),
                      J = length(players),
                      y = final_data$focal_result,
                      focal_id = final_data$focal_id,
                      opp_id = final_data$opp_id,
                      focal_prev = final_data$focal_prev_result,
                      opp_prev = final_data$opp_prev_result,
                      focal_avg = final_data$focal_avg,
                      opp_avg = final_data$opp_avg)

@@

>>>>

library(tidyverse)
library(RcppRoll)
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(here)

@@

get_hist <- function(user, games, prev_n) {
  hist_games <- games %>% 
    filter(White == user | Black == user) %>% 
    arrange(UTCDate, UTCTime) %>% 
    mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
    select(White:BlackElo, focal_white, game_id) %>% 
    mutate(focal_result = case_when(
      (focal_white == 1 & Result == "1-0") ~ 1,
      (focal_white == 0 & Result == "0-1") ~ 1,
      (Result == "1/2-1/2") ~ 0.5,
      .default = 0
      )) %>% 
    mutate(focal_win_prop = c(cumsum(focal_result[1:(prev_n - 1)])/(1:(prev_n -1)), 
                              roll_mean(focal_result, n = prev_n)))
  hist_games
}

@@

## then just get all the history for this
all_hist <- map_dfr(players, get_hist_opp, bullet_60, prev_n = 5) %>% 
  as_tibble() %>% 
  select(game_id:player, UTCDate, UTCTime)

min_hist <- all_hist %>% select(game_id:player)

@@

tidy_stan <- tidy_bullet %>% 
  left_join(min_hist, by = c("game_id", "Username" = "player")) %>% 
  arrange(UTCDate, UTCTime) %>% 
  rename(focal_result = opp_result, focal_win_prop = opp_win_prop) %>% 
  left_join(min_hist, by = c("game_id", "opponent" = "player")) %>% 
  as_tibble()

@@

## getting the overall current win probabilities for each player also, 
## up to and including the current game

player_avg <- all_hist %>% 
  group_by(player) %>% 
  arrange(UTCDate, UTCTime, .by_group = TRUE) %>%
  mutate(nrow = row_number()) %>% 
  mutate(curr_avg = cumsum(opp_result)/nrow) %>% 
  select(game_id, player, curr_avg)

global_avg <- all_hist %>% 
  group_by(player) %>% 
  summarise(total_avg = mean(opp_result))

@@

final_data <- tidy_stan %>% 
  left_join(player_avg, by = c("Username" = "player", "game_id")) %>% 
  rename(focal_avg = curr_avg) %>% 
  left_join(player_avg, by = c("opponent" = "player", "game_id")) %>% 
  rename(opp_avg = curr_avg) %>% 
  group_by(Username) %>% 
  arrange(UTCDate, UTCTime) %>% 
  mutate(focal_prev_result = lag(focal_result, default = 0),
         focal_runn_avg = lag(focal_avg, n = 2, default = 0)) %>% 
  ungroup() %>% 
  group_by(opponent) %>% 
  arrange(UTCDate, UTCTime) %>% 
  mutate(opp_prev_result = lag(opp_result, default = 0),
         opp_runn_avg = lag(opp_avg, n = 2, default = 0)) %>% 
  ungroup() %>% 
  mutate(focal_id = match(Username, players),
         opp_id = match(opponent, players)) %>% 
  filter(focal_result != 0.5) %>% 
  select(-focal_avg, - opp_avg, - opp_result, 
         - focal_win_prop, - opp_win_prop)

final_data
## remove some unused columns

norm_data <- final_data %>% 
  left_join(global_avg, by = c("Username" = "player")) %>% 
  rename(focal_global_avg = total_avg) %>% 
  left_join(global_avg, by = c("opponent" = "player")) %>% 
  rename(opp_global_avg = total_avg) %>% 
  mutate(focal_prev_result = focal_prev_result - focal_global_avg,
         focal_runn_avg = focal_runn_avg - focal_global_avg,
         opp_prev_result = opp_prev_result - opp_global_avg,
         opp_runn_avg = opp_runn_avg - opp_global_avg)

@@

Now this matches nicely, contains their historic winning average up to 2 games
previously, along with their result in the previous game.

Note that this has not been standardized currently.

@@

norm_data

stan_data_all <- list(N = nrow(norm_data),
                      J = length(players),
                      y = norm_data$focal_result,
                      focal_id = norm_data$focal_id,
                      opp_id = norm_data$opp_id,
                      focal_prev = norm_data$focal_prev_result,
                      opp_prev = norm_data$opp_prev_result,
                      focal_avg = norm_data$focal_runn_avg,
                      opp_avg = norm_data$opp_runn_avg)

@@

## Fitting this to all data

Fit this model to all data on the cluster, just to see the results.
The traceplots and R-hat for this model indicates it doesn't work well,
does not converge.

## Comparing models on smaller data

Is there a small subset of players who play each other often that we can
use in the dataset that can be used to compare these models?

bullet_60 %>% 
  filter(White %in% focal_players) %>% 
  filter(Black %in% focal_players)

## no games between any of the players used to scrape the data
## would maybe need to get more players then, to use this for a comparison

```{r}
library(chessR)

chessdotcom_leaders <- chessdotcom_leaderboard(game_type = "live_bullet")

top_bullet <- chessdotcom_leaders %>% top_n(10, wt = -rank) %>% 
  select(player_id, username)

chessdotcom_hikaru_recent <- get_raw_chessdotcom(usernames = "Hikaru",
                                                 year_month = c(202201:202205))

jan_games <- top_bullet$username %>% 
  map_dfr(~get_raw_chessdotcom(.x, year_month = 202201))

top_games <- jan_games %>% 
  filter(White %in% top_bullet$username) %>% 
  filter(Black %in% top_bullet$username)

## get 500 games here then for 1 month, probably enough to fit an initial model
## most of these games bullet too, so could use that

top_games %>% 
  group_by(White, Black, TimeClass) %>% 
  tally() %>% 
  arrange(-n)

top_games %>% 
  filter(TimeClass == "blitz")
